Выпускная квалификационная работа

### 2025-06-02 TeX/CDAM/en_cdam_KhromovGolyandina.tex
add refs

  \abskeyw{Tensor modifications of singular spectrum analysis for  signal extraction and frequency estimation problems in a noisy sum  of exponentially modulated sinusoids are reviewed. Modifications using Higher-Order SVD are considered. Numerical comparisons are  carried out. It is shown numerically that for, the signal extraction problem, tensor methods generally perform worse than matrix methods for a single-channel series, but can outperform multichannel SSA for a series system. For frequency estimation, tensor modifications are generally advantageous.}{time series,
  signal, frequency estimation, tensor, singular spectrum analysis}
\end{center}

@@

Singular spectrum analysis (SSA) is one of the methods used for time series analysis~\cite{Golyandina2001}, in which the original time series is transformed into a matrix, called the trajectory matrix, using a given window length $L$. The singular value decomposition (SVD) of this matrix is then analyzed. When the objective is to estimate the signal and its properties from an observed noisy series, the first $r$ components of the SVD are considered, where $r$ is the rank of the signal trajectory matrix. Based on the selected components, the signal estimation is constructed. A distinctive feature of the method is that it does not require the specification of a signal model. However, SSA can also handle a parametric signal model in the form of a sum of products of polynomials, exponentials and sinusoids. The frequency estimation problem plays a special role. The ESPRIT method uses the estimation of the signal subspace based on the $r$ leading left singular vectors of the trajectory matrix SVD to estimate the frequencies present in the signal.
The least squares (LS) version of ESPRIT~\cite{Roy1986} is also known as Hankel SVD (HSVD), and the total least squares (TLS) version is known as HTLS \cite{https://www.sciencedirect.com/science/article/abs/pii/S1064185884712095}.
%Пусть сигнал задан в виде суммы синусоид (или комплексных экспонент
% в комплексном случае).
%Свойства оценок частот и оценок сигнала сильно различаются. В
% частности, дисперсия  оценки сигнала имеет порядок $1/N$, в то
% время как дисперсия оценок частот имеет порядок $1/N^3$, где $N$
% --- длина временного ряда и шум белый, гауссовский.

A number of works propose tensor modifications of the SSA and ESPRIT
methods, where the original series is transformed into a tensor,
usually of 3rd order, instead of a matrix \cite{at least three refs}.  One of the common
variants of tensor decompositions is Higher-Order SVD (HO-SVD), which
generalizes the matrix SVD.

@@

Let $\tX = (\tX^{(1)}, \tX^{(2)}, \ldots, \tX^{(P)})$ be a multi-channel
time series consisting of $P$ single-channel series, also called channels.
\begin{definition}
  The tensor embedding operator for a multi-channel time series with
  window length $L$ (that is, $I=P$, $L$, $K=N-L+1$) such that ${1< L < N}$ is a mapping $\calT_{L}$
  that translates the
  $P$-channel time-series $\tX$ into the tensor $\calX \in
  \mathbb{C}^{L\times K \times P}$ (${K = N - L + 1}$) as follows
  $\mathcal{X}_{lkp} =  x_{l+k-1}^{(p)}$, where ${l \in
  \overline{1:L}},\, {k \in \overline{1:K}},\, {p \in \overline{1:P}}$.
\end{definition}

@@

\subsection{Dstack modifications}
In the paper~\cite{Papy2009}, to improve the speed of the method, it
is proposed to transform a single-channel series into a multi-channel
series before applying the tensor modification: $x_m^{(d)} =
x_{(m-1)D + d}$, where $m \in \overline{1:(N/D)}$. In that paper only
the ESPRIT modification called HTLSDstack is considered, but we will
apply this time series transformation for the signal estimation
problem as well, and will call the resulting method SSADstack.
Tensor modifications are constructed as for a multichannel series.

@@

\begin{thebibliography}{10}
  \bibitem{Golyandina2001}
  Golyandina~N.E., Nekrutkin~V.V., Zhigljavsky~A.A.~(2001).
  {\sl Analysis of Time Series Structure}.
  Chapman and Hall/CRC: Boca Raton.

  \bibitem{Roy1986}
  Roy~R., Paulraj~A., Kailath~T.~(1986).
  ESPRIT-A subspace rotation approach to estimation of parameters of
  cisoids in noise.
  {\sl IEEE Transactions on Acoustics, Speech, and Signal Processing}.
  Vol.~{\bf 34}, Num.~{\bf 5},
  pp.~1340-1342.

@@

  % \bibitem{paper}
  % Jacobs~P.A., Lewis~P.A.W.~(1983). Stationary Discrete
  % Autoregressive-Moving Average
  % Time Series Generated by Mixtures. {\sl Journal of Time Series
  % Analysis}. Vol.~{\bf 4}, Num.~{\bf 1},
  % pp.~19-36.

  % \bibitem{book}
  % Johnson~N.L., Kotz~S., Balakrishnan~N.~(1997). {\sl Discrete
  % Multivariate Distributions}. Wiley: New York.

>>>>

  \abskeyw{Tensor modifications of singular spectrum analysis for  signal extraction and frequency estimation problems in a noisy sum  of exponentially modulated sinusoids are reviewed. Modifications using Higher-Order SVD are considered. Numerical comparisons are  carried out. It is shown numerically that for, the signal extraction problem, tensor methods generally perform worse than matrix methods for a single-channel series, but can outperform multi-channel SSA for a series system. For frequency estimation, tensor modifications are generally advantageous.}{time series,
  signal, frequency estimation, tensor, singular spectrum analysis}
\end{center}

@@

Singular spectrum analysis (SSA) is one of the methods used for time series analysis~\cite{Golyandina2001}, in which the original time series is transformed into a matrix, called the trajectory matrix, using a given window length $L$. The singular value decomposition (SVD) of this matrix is then analyzed. When the objective is to estimate the signal and its properties from an observed noisy series, the first $r$ components of the SVD are considered, where $r$ is the rank of the signal trajectory matrix. Based on the selected components, the signal estimation is constructed. A distinctive feature of the method is that it does not require the specification of a signal model. However, SSA can also handle a parametric signal model in the form of a sum of products of polynomials, exponentials and sinusoids. The frequency estimation problem plays a special role. The ESPRIT method uses the estimation of the signal subspace based on the $r$ leading left singular vectors of the trajectory matrix SVD to estimate the frequencies present in the signal.
The least squares (LS) version of ESPRIT~\cite{Roy1989} is also known as Hankel SVD (HSVD), and the total least squares (TLS) version is known as HTLS \cite{VanHuffel1994}.
%Пусть сигнал задан в виде суммы синусоид (или комплексных экспонент
% в комплексном случае).
%Свойства оценок частот и оценок сигнала сильно различаются. В
% частности, дисперсия  оценки сигнала имеет порядок $1/N$, в то
% время как дисперсия оценок частот имеет порядок $1/N^3$, где $N$
% --- длина временного ряда и шум белый, гауссовский.

A number of works propose tensor modifications of the SSA and ESPRIT
methods, where the original series is transformed into a tensor,
usually of 3rd order, instead of a matrix \cite{DeLathauwer2011,Papy2009,Trung2024}.  One of the common
variants of tensor decompositions is Higher-Order SVD (HO-SVD), which
generalizes the matrix SVD.

@@

Let $\tX = (\tX^{(1)}, \tX^{(2)}, \ldots, \tX^{(P)})$ be a multi-channel
time series consisting of $P$ single-channel series, also called channels.
\begin{definition}
  The tensor embedding operator for a multi-channel time series with
  window length $L$ (that is, $I=P$, $L$, $K=N-L+1$) such that ${1< L < N}$ is a mapping $\calT_{L}$
  that translates the
  $P$-channel time-series $\tX$ into the tensor $\calX \in
  \mathbb{C}^{L\times K \times P}$ as follows
  $\mathcal{X}_{lkp} =  x_{l+k-1}^{(p)}$, where ${l \in
  \overline{1:L}},\, {k \in \overline{1:K}},\, {p \in \overline{1:P}}$.
\end{definition}

@@

\subsection{Dstack modifications}
In the paper~\cite{Papy2009}, to improve the speed of the method, it
is proposed to transform a single-channel series into a multi-channel
series before applying the tensor modification: $x_m^{(d)} =
x_{(m-1)D + d}$, where $m \in \overline{1:(N/D)}$. In that paper only
the ESPRIT modification called HTLSDstack is considered, but we will
apply this time series transformation for the signal estimation
problem as well, and will call the resulting method SSADstack.
Tensor modifications are constructed as for a multi-channel series.

@@

\begin{thebibliography}{10}
  \bibitem{DeLathauwer2011}
  De Lathauwer~L.~(2011). Blind separation of exponential polynomials and the decomposition of a tensor in rank-$(L\_r,L\_r,1)$ terms. {\sl SIAM Journal on Matrix Analysis and Applications}. Vol.~{\bf 32}, Num.~{\bf 4},
  pp.~1451-1474.

  \bibitem{Golyandina2001}
  Golyandina~N.E., Nekrutkin~V.V., Zhigljavsky~A.A.~(2001).
  {\sl Analysis of Time Series Structure}.
  Chapman and Hall/CRC: Boca Raton.

@@

  \bibitem{Trung2024}
  Trung Le~T., Abed-Meraim~K., Trung ~N.L.~(2024). Higher-Order Singular Spectrum Analysis for Multichannel Biomedical Signal Analysis. {\sl 2024 32nd European Signal Processing Conference (EUSIPCO), Lyon, France}. Pp.~1337-1341.

  \bibitem{VanHuffel1994}
  Van Huffel~S., Chen~H.,Decanniere~C., van Hecke~P.~(1994). Algorithm for time-domain {NMR} data fitting based on total least squares. {\sl Journal of Magnetic Resonance, Series A}. Vol.~{\bf 110}, Num.~{\bf 2},
  pp.~19-36.
